\Large{Adversarial Training against Systematic Uncertainty}

### 2016-10-20 tex/paper.tex
simpler notation for entropy

In this setting, the nuisance parameter $Z$ is considered as a random variable
of bounded support, for which we require the uniform prior $p(z)$ (for $z \in
{\cal Z}$). Importantly, classification of $Y$ with respect to $X$ is therefore
considered in the context where $Z$ is marginalized out, which means that the
classifier minimizing ${\cal L}_f$ is optimal with respect to $Y|X$, but not
necessarily with $Y|X,Z$. Results hold for a nuisance parameter $Z$ taking
either categorical values or continuous values within a bounded support. By
abuse of notation, $H(p_Z)$ denotes the differential entropy in this latter
case. Finally, propositions below are derived in a non-parametric setting, by
assuming that both $f$ and $r$ have enough capacity.

@@

\begin{proposition}\label{prop:2}
If there exists a saddle point $(\smash{\hat\theta}_f, \smash{\hat\theta}_r)$
for Eqn.~\ref{eqn:min_thetaf} and \ref{eqn:max_thetar} such that
$E(\hat\theta_f, \hat\theta_r) = \mathbb{E}_{X} [H(p_{Y|X}) - H(p_Z)]$, then
$f(\cdot;\smash{\hat\theta}_f)$ is both an optimal classifier and a pivotal
quantity.
\end{proposition}

@@

For fixed $\theta_f$, the adversary $r$ is optimal at $\hat\theta_r = \arg
\max_{\theta_r} E(\theta_f, \theta_r)  = \arg \min_{\theta_r} {\cal
L}_r(\theta_f, \theta_r)$, in which case $p_{\hat\theta_r}(z|f(X;\theta_f)) =
p(z|f(X;\theta_f))$ and ${\cal L}_r$ reduces to the expected entropy
$\mathbb{E}_{X} [ H(p_{Z|f(X;\theta_f)}) ]$ of the conditional distribution of the nuisance. The
value function $E$ can therefore be rewritten as $$E'(\theta_f) = {\cal
L}_f(\theta_f) - \mathbb{E}_{X} [ H(p_{Z|f(X;\theta_f)}) ].$$  In particular, we have the lower
bound $$\mathbb{E}_{X} [ H(p_{Y|X}) - H(p_Z)] \leq {\cal L}_f(\theta_f) - \mathbb{E}_{X} [H(p_{Z|f(X;\theta_f)})]$$
where the equality holds at $\smash{\hat\theta}_f = \arg \min_{\theta_f}
E'(\theta_f)$ only when
\begin{itemize}
    \item $\smash{\hat\theta}_f$ minimizes the negative log-likelihood ${\cal
    L}_f$ of $X|Y$, which happens when $\smash{\hat\theta}_f$ are the parameters
    of an optimal classifier and in which case ${\cal L}_f$ reduces to its
    minimum value $\mathbb{E}_{X} [H(p_{Y|X})]$,

    \item $\smash{\hat\theta}_f$ maximizes the expected entropy
    $\mathbb{E}_{X} [H(p_{Z|f(X;\theta_f)})]$, which happens when all outcomes of
    $Z|f(X;\smash{\hat\theta}_f)$ are equally likely. In other words,
    $p(z|f(X;\smash{\hat\theta}_f)) = p(z)$ for all $z \in {\cal Z}$ and all values of $f(X;\theta_f)$ since we
    require a uniform prior by construction. Note that in the continuous case,
    the supremum of the differential entropy over continuous distributions on
    the same bounded support is also realized by the uniform distribution over
    that support.
\end{itemize}
When the lower bound is active, we have $p_{\theta_r}(z|f(X;\theta_f)) = p(z)$
because of the second condition, which induces that the
optimal classifier $f(\cdot;\smash{\hat\theta}_f)$ is also a pivotal
quantity as a consequence of Proposition~\ref{prop:1}.

@@

Proposition~\ref{prop:2} suggests that if at each step of
Algorithm~\ref{alg:adversarial-training} the adversary $r$ is allowed to reach
its optimum given $f$ (e.g., by setting $K$ sufficiently high) and if $f$ is
updated to improve ${\cal L}_f(\theta_f) - \mathbb{E}_X [ H(p_{Z|f(X;\theta_f)}) ]$, then $f$
should converge to a classifier which is both optimal and pivotal, provided such
a classifier exists. On many problems of interest though, such a classifier may
not exist because the nuisance parameter directly shapes the decision boundary,
in which cases the lower bound $\mathbb{E}_{X} [ H(p_{Y|X}) - H(p_Z)] < {\cal L}_f(\theta_f) - \mathbb{E}_{X} [H(p_{Z|f(X;\theta_f)})]$ is strict: $f$ can either be an optimal classifier or a
pivotal quantity, but not both simultaneously. In this situation, it is natural
to rewrite the value function $E$  as
\begin{equation}
    E_\lambda(\theta_f, \theta_r) = {\cal L}_f(\theta_f) - \lambda {\cal L}_r(\theta_f, \theta_r),
\end{equation}
where $\lambda \geq 0$ is a hyper-parameter controlling the trade-off between
the performance of $f$ and its independence with respect to the nuisance
parameter. Setting $\lambda$ to a large value will preferably enforces $f$ to
be pivotal while setting $\lambda$ close to $0$ will rather constraint $f$ to be
optimal.

@@

Let us now consider the adversarial training of $f$ against an adversary $r$
implemented as a mixture density network modeling $Z|f(X;\theta_f)$ as a
 mixture of five gaussians. As for $f$, the network
architecture comprises 2 dense hidden layers of 20 nodes with ReLU activations,
but is followed by an output layer of 15 nodes corresponding to the means,
standard deviations and mixture coefficients of the five gaussians. Output nodes
for the mean values come with linear activations, nodes
for the standard deviations with exponential activations to ensure positivity,
while nodes for the mixture coefficients implement the softmax function to
ensure positivity and normalization. When running
Algorithm~\ref{alg:adversarial-training} as initialized with the classifier $f$
obtained previously, adversarial training effectively reshapes the decision
function so it that becomes almost independent on the nuisance parameter, as
shown in the lower plots of Figure~\ref{fig:toy}. In particular, the conditional
probability densities of the decision scores $f(X;\theta_f)$ are now very
similar to each other, indicating only a small  dependency on the nuisance, as
theoretically expected. The dynamics of adversarial training is illustrated in
Figure~\ref{fig:toy-training}, where the losses ${\cal L}_f$, ${\cal L}_r$ and
${\cal L}_f - \lambda {\cal L}_r$ are evaluated after each iteration of
Algorithm~\ref{alg:adversarial-training}. In the first iterations, we observe
that the global objective ${\cal L}_f - \lambda {\cal L}_r$ is minimized by
making the classifier less accurate, hence the corresponding increase of ${\cal
L}_f$, but which results in a classifier that is more pivotal, hence the
corresponding increase of ${\cal L}_r$ and the total net benefit. As learning
goes, minimizing $E$ then requires making predictions that are more accurate,
hence decreasing ${\cal L}_f$, or that are less dependent on $Z$, hence shaping
$p_{\theta_r}$ so that it becomes more uniform. Indeed, ${\cal L}_f$ eventually
start to slightly decrease, while remaining lower bounded by $\min_{\theta_f}
{\cal L}_f(\theta_f)$ as approximated by the dashed line in the first plot.
Similarly,  ${\cal L}_r$ keeps tending towards the differential entropy $H(p_Z) =
\log(\sigma \sqrt{2 \pi e}) = 1.419$ of the gaussian distribution with unit variance, as shown by the dashed
line in the second plot. Finally, let us note that the ideal situation of a
classifier that is both optimal and pivotal appears to be unreachable for this
problem, as shown in third plot by the  offset between ${\cal L}_f - \lambda
{\cal L}_r$ and the dashed line approximating $\mathbb{E}_{X} [ H(p_{Y|X}) -
\lambda H(p_Z)]$.

>>>>

In this setting, the nuisance parameter $Z$ is considered as a random variable
of bounded support, for which we require the uniform prior $p(z)$ (for $z \in
{\cal Z}$). Importantly, classification of $Y$ with respect to $X$ is therefore
considered in the context where $Z$ is marginalized out, which means that the
classifier minimizing ${\cal L}_f$ is optimal with respect to $Y|X$, but not
necessarily with $Y|X,Z$. Results hold for a nuisance parameter $Z$ taking
either categorical values or continuous values within a bounded support. By
abuse of notation, $H(Z)$ denotes the differential entropy in this latter
case. Finally, propositions below are derived in a non-parametric setting, by
assuming that both $f$ and $r$ have enough capacity.

@@

\begin{proposition}\label{prop:2}
If there exists a saddle point $(\smash{\hat\theta}_f, \smash{\hat\theta}_r)$
for Eqn.~\ref{eqn:min_thetaf} and \ref{eqn:max_thetar} such that
$E(\hat\theta_f, \hat\theta_r) = \mathbb{E}_{X} [H({Y|X}) - H(Z)]$, then
$f(\cdot;\smash{\hat\theta}_f)$ is both an optimal classifier and a pivotal
quantity.
\end{proposition}

@@

For fixed $\theta_f$, the adversary $r$ is optimal at $\hat\theta_r = \arg
\max_{\theta_r} E(\theta_f, \theta_r)  = \arg \min_{\theta_r} {\cal
L}_r(\theta_f, \theta_r)$, in which case $p_{\hat\theta_r}(z|f(X;\theta_f)) =
p(z|f(X;\theta_f))$ and ${\cal L}_r$ reduces to the expected entropy
$\mathbb{E}_{X} [ H({Z|f(X;\theta_f)}) ]$ of the conditional distribution of the nuisance. The
value function $E$ can therefore be rewritten as $$E'(\theta_f) = {\cal
L}_f(\theta_f) - \mathbb{E}_{X} [ H({Z|f(X;\theta_f)}) ].$$  In particular, we have the lower
bound $$\mathbb{E}_{X} [ H({Y|X}) - H(Z)] \leq {\cal L}_f(\theta_f) - \mathbb{E}_{X} [H({Z|f(X;\theta_f)})]$$
where the equality holds at $\smash{\hat\theta}_f = \arg \min_{\theta_f}
E'(\theta_f)$ only when
\begin{itemize}
    \item $\smash{\hat\theta}_f$ minimizes the negative log-likelihood ${\cal
    L}_f$ of $X|Y$, which happens when $\smash{\hat\theta}_f$ are the parameters
    of an optimal classifier and in which case ${\cal L}_f$ reduces to its
    minimum value $\mathbb{E}_{X} [H({Y|X})]$,

    \item $\smash{\hat\theta}_f$ maximizes the expected entropy
    $\mathbb{E}_{X} [H({Z|f(X;\theta_f)})]$, which happens when all outcomes of
    $Z|f(X;\smash{\hat\theta}_f)$ are equally likely. In other words,
    $p(z|f(X;\smash{\hat\theta}_f)) = p(z)$ for all $z \in {\cal Z}$ and all values of $f(X;\theta_f)$ since we
    require a uniform prior by construction. Note that in the continuous case,
    the supremum of the differential entropy over continuous distributions on
    the same bounded support is also realized by the uniform distribution over
    that support.
\end{itemize}
When the lower bound is active, we have $p_{\theta_r}(z|f(X;\theta_f)) = p(z)$
because of the second condition, which induces that the
optimal classifier $f(\cdot;\smash{\hat\theta}_f)$ is also a pivotal
quantity as a consequence of Proposition~\ref{prop:1}.

@@

Proposition~\ref{prop:2} suggests that if at each step of
Algorithm~\ref{alg:adversarial-training} the adversary $r$ is allowed to reach
its optimum given $f$ (e.g., by setting $K$ sufficiently high) and if $f$ is
updated to improve ${\cal L}_f(\theta_f) - \mathbb{E}_X [ H({Z|f(X;\theta_f)}) ]$, then $f$
should converge to a classifier which is both optimal and pivotal, provided such
a classifier exists. On many problems of interest though, such a classifier may
not exist because the nuisance parameter directly shapes the decision boundary,
in which cases the lower bound $\mathbb{E}_{X} [ H({Y|X}) - H(Z)] < {\cal L}_f(\theta_f) - \mathbb{E}_{X} [H({Z|f(X;\theta_f)})]$ is strict: $f$ can either be an optimal classifier or a
pivotal quantity, but not both simultaneously. In this situation, it is natural
to rewrite the value function $E$  as
\begin{equation}
    E_\lambda(\theta_f, \theta_r) = {\cal L}_f(\theta_f) - \lambda {\cal L}_r(\theta_f, \theta_r),
\end{equation}
where $\lambda \geq 0$ is a hyper-parameter controlling the trade-off between
the performance of $f$ and its independence with respect to the nuisance
parameter. Setting $\lambda$ to a large value will preferably enforces $f$ to
be pivotal while setting $\lambda$ close to $0$ will rather constraint $f$ to be
optimal.

@@

Let us now consider the adversarial training of $f$ against an adversary $r$
implemented as a mixture density network modeling $Z|f(X;\theta_f)$ as a
 mixture of five gaussians. As for $f$, the network
architecture comprises 2 dense hidden layers of 20 nodes with ReLU activations,
but is followed by an output layer of 15 nodes corresponding to the means,
standard deviations and mixture coefficients of the five gaussians. Output nodes
for the mean values come with linear activations, nodes
for the standard deviations with exponential activations to ensure positivity,
while nodes for the mixture coefficients implement the softmax function to
ensure positivity and normalization. When running
Algorithm~\ref{alg:adversarial-training} as initialized with the classifier $f$
obtained previously, adversarial training effectively reshapes the decision
function so it that becomes almost independent on the nuisance parameter, as
shown in the lower plots of Figure~\ref{fig:toy}. In particular, the conditional
probability densities of the decision scores $f(X;\theta_f)$ are now very
similar to each other, indicating only a small  dependency on the nuisance, as
theoretically expected. The dynamics of adversarial training is illustrated in
Figure~\ref{fig:toy-training}, where the losses ${\cal L}_f$, ${\cal L}_r$ and
${\cal L}_f - \lambda {\cal L}_r$ are evaluated after each iteration of
Algorithm~\ref{alg:adversarial-training}. In the first iterations, we observe
that the global objective ${\cal L}_f - \lambda {\cal L}_r$ is minimized by
making the classifier less accurate, hence the corresponding increase of ${\cal
L}_f$, but which results in a classifier that is more pivotal, hence the
corresponding increase of ${\cal L}_r$ and the total net benefit. As learning
goes, minimizing $E$ then requires making predictions that are more accurate,
hence decreasing ${\cal L}_f$, or that are less dependent on $Z$, hence shaping
$p_{\theta_r}$ so that it becomes more uniform. Indeed, ${\cal L}_f$ eventually
start to slightly decrease, while remaining lower bounded by $\min_{\theta_f}
{\cal L}_f(\theta_f)$ as approximated by the dashed line in the first plot.
Similarly,  ${\cal L}_r$ keeps tending towards the differential entropy $H(Z) =
\log(\sigma \sqrt{2 \pi e}) = 1.419$ of the gaussian distribution with unit variance, as shown by the dashed
line in the second plot. Finally, let us note that the ideal situation of a
classifier that is both optimal and pivotal appears to be unreachable for this
problem, as shown in third plot by the  offset between ${\cal L}_f - \lambda
{\cal L}_r$ and the dashed line approximating $\mathbb{E}_{X} [ H({Y|X}) -
\lambda H(Z)]$.
